Evidence of off-target probe binding in the 10x Genomics Xenium v1 Human Breast Gene Expression Panel compromises accuracy of spatial transcriptomic profiling

Version 1

Abstract

Abstract

The accuracy of spatial gene expression profiles generated by probe-based in situ spatially-resolved transcriptomic technologies depends on the specificity with which probes bind to their intended target gene. Off-target binding, defined as a probe binding to something other than the target gene, can distort a gene’s true expression profile, making probe specificity essential for reliable transcriptomics. Here, we investigate off-target binding in the 10x Genomics Xenium v1 Human Breast Gene Expression Panel. We developed a software tool, Off-target Probe Tracker (OPT), to identify putative off-target binding via alignment of probe sequences and found at least 21 out of the 280 genes in the panel impacted by off-target binding to protein-coding genes. To substantiate our predictions, we leveraged a previously published Xenium breast cancer dataset generated using this gene panel and compared results to orthogonal spatial and single-cell transcriptomic profiles from Visium CytAssist and 3ʹ single-cell RNA-seq derived from the same tumor block. Our findings indicate that for some genes, the expression patterns detected by Xenium demonstrably reflect the aggregate expression of the target and predicted off-target genes based on Visium and single-cell RNA-seq rather than the target gene alone. Overall, this work enhances the biological interpretability of spatial transcriptomics data and improves reproducibility in spatial transcriptomics research.

Introduction

Recent advances in high-throughput spatially-resolved transcriptomic profiling technologies have enabled the investigation of gene expression with high spatial resolution within tissues. One such commercially available spatial transcriptomics platform is Xenium from 10x Genomics, a publicly traded company with a market capitalization exceeding $1 billion as of March 2025 (Yahoo Finance 2025). Xenium achieves spatial gene expression profiling at single-cell resolution for targeted genes using a probe-based in situ detection approach. 10x Genomics currently offers targeted gene panels with pre-designed probe sets, including their 280-gene Xenium v1 Human Breast Gene Expression Panel. As of December 2024, over 16,000 Xenium consumable reactions have been sold, with each tissue slide profiled costing approximately $5,000, underscoring the platform’s widespread use and high commercial value (10x Genomics).

Briefly, Xenium uses padlock probes that include sequences complementary to the RNA of target genes. Once a padlock probe binds to its target, it is ligated and subsequently amplified via rolling circle amplification (RCA). Fluorescently labeled decoder probes then hybridize to the amplified RCA product, enabling the simultaneous detection and decoding of the optical signature, or codeword, specific to each gene in the panel through successive rounds of fluorescence imaging. When combined with cell-segmentation, this approach allows for spatially-resolved single-cell quantification of gene expression.

The accuracy of these gene expression measurements thus relies on the specificity of the probes to bind to their intended target gene. We define off-target binding as when a probe binds to something other than the RNA sequence intended to correspond to the target gene (Figure 1). We note once ligation and RCA occur, the resulting fluorescent signal cannot be distinguished between on-target and off-target binding. As such, off-target binding can distort the quantification of the intended target gene’s expression, as the observed expression would represent a combination of the target as well as off-target expression.

Schematic of potential off-target binding in 10x Xenium.

In this illustration, the arms of the padlock probes were designed to bind an RNA sequence intended to correspond to a target gene (green). However, these probes exhibit off-target binding and bind to an RNA sequence in a different off-target gene (red). The probe is circularized and subsequently amplified via rolling circle amplification (RCA). Hybridization of fluorescent probes to the RCA product enables the generation a fluorescent signal that is used to quantify RNA expression within cells.

To predict for such potential off-target binding, we developed Off-target Probe Tracker (OPT), a software tool that aligns probe sequences to an annotated transcriptome with the option to allow for mismatches that may still permit probe binding. Using OPT, we identify putative off-target probe binding to protein-coding genes affecting at least 21 out of 280 genes in the 10x Genomics Xenium v1 Human Breast Gene Expression Panel, compromising the accuracy of their spatial transcriptomic profiles. We substantiate our predictions using data from orthogonal spatial and single-cell gene expression profiling technologies. By facilitating a more rigorous evaluation of probe sequence specificity, tools like OPT can aid in future probe design decisions to help ensure that probes are optimized to minimize off-target binding based on current transcriptome annotations.

Results

OPT predicts potential off-target probe binding

To identify potential off-target binding impacting the 10x Genomics Xenium v1 Human Breast Gene Expression Panel, we first downloaded the FASTA file containing the probe sequences from the 10x Genomics Website (Methods, Supplementary Data). This file includes 4,809 probe sequences that are 40bp in length and designed to target 280 genes, averaging 17 probe sequences per gene (ranging from 2 to 50 probe sequences per gene).

We developed a software tool called OPT (Methods) that uses nucmer (Marçais et al. 2018) to predict off-target binding by aligning probe sequences to a reference transcriptome, which comprises a curated collection of transcript isoforms for all genes in a species. OPT features adjustable parameters for binding strictness (e.g., number of mismatches) and generates a summary file that details all targeted genes along with their potential off-targets based on the sequence alignments. 10x Genomics designed its probe sequences using the GENCODE “basic” annotation (Mudge et al. 2025), so initially we also used the latest GENCODE “basic” annotation (v47) to predict off-target binding for these probes.

We first sought to predict if a probe has off-target binding based on perfect sequence homology (i.e., if it aligns with 100% identity) with any annotated transcripts other than those that belong to the intended target gene. Of the 4,809 probe sequences in this gene panel, using GENCODE v47, OPT identified 180 probe sequences across 45 genes as having off-target binding based on perfect sequence homology (Table 1). Among the 45 genes with predicted off-target binding, the number of affected probe sequences per gene ranged from 1 to 14. Overall, these off-target probes matched 90 other genes, including 34 protein-coding genes, 34 pseudogenes, 12 long non-coding RNA, 9 transcripts labeled as nonsense-mediated decay, and one microRNA gene.

Variation in off-target binding predictions across different annotations

OPT relies on alignments to an annotated transcriptome, which ideally reflects all genes and gene variants stably transcribed in a given species. However, genome annotation is still an active area of research (Varabyou et al. 2023), with discrepancies across annotations in gene counts, isoforms, and many other features. We therefore further used OPT to predict for off-target binding and affected genes using two additional human genome annotation sets, RefSeq (v110), (O’Leary et al. 2016) and CHESS (v3.1.3) (Varabyou et al. 2023), and compared to our previous results from GENCODE (v47) (Mudge et al. 2025) (Methods).

When considering only perfect sequence homology, while we previously found 45 affected genes using GENCODE, we found 22 when using RefSeq and 33 when using CHESS (Supplementary Table 1, 2). Given that RefSeq and CHESS have more transcripts than GENCODE, these discrepancies in off-target binding predictions was not simply an artifact of the difference in transcript set sizes.

While the human annotation databases mostly agree on the number of protein-coding genes in the genome, they remain widely divergent on pseudogenes and lncRNA genes. Therefore, we focused on how the results change when we restrict our analysis to only protein-coding genes. By excluding pseudogenes (which are presumably not expressed), lncRNAs, and other non-protein-coding RNAs when using OPT, the number of affected genes fell to 17 for GENCODE and RefSeq and 16 for CHESS (Supplementary Table 3). Again, these discrepancies reflect annotation differences. For example, the probe sequence (ENSG00000196154|S100A4|ab4e3dc), which was designed to target S100A4 based on GENCODE annotations, also aligns to S100A5 in RefSeq (Supplementary Figure 1A). We reason that if a probe sequence aligns off-target to a protein-coding gene based on any of these annotations, it could result in off-target binding. We therefore focused further analysis on the union of genes with predicted off-target binding to protein-coding genes across the 3 annotations, resulting in 21 genes: ADGRE5, ADH1B, AKR1C1, AKR1C3, APOBEC3A, APOBEC3B, AQP1, C1QA, CD79B, CEACAM6, FCGR3A, FLNB, KLRC1, KRT6B, PDGFRA, S100A4, TOMM7, TPD52, TPSAB1, TUBA4A, TUBB2B (Supplementary Table 4).

Comparison with Visium CytAssist reveals spatial gene expression patterns consistent with off-target binding

To investigate the potential effects of our predicted off-target binding for the Xenium v1 Human Breast Gene Expression Panel in experimental settings, we compared spatial gene expression patterns detected in two previously published spatial transcriptomics datasets from serial sections of the same breast cancer tissue: one section assayed with Xenium using the Xenium v1 Human Breast Gene Expression Panel, and another assayed using Visium CytAssist, an orthogonal spatial transcriptomics platform (Janesick et al. 2023). Briefly, Visium is a sequencing-based spatial transcriptomics technology in which RNA is hybridized to spatially barcoded capture spots on a slide, enabling spatial transcriptomic mapping after sequencing. However, while Xenium offers single-cell resolution gene expression quantification, Visium quantifies gene expression within 55μm spots. To enable direct comparison, we first structurally aligned the Xenium and Visium tissue sections using STalign (Clifton et al. 2023), restricting our analysis to overlapping regions since different parts of the tissue were profiled (Supplementary Figure 2A). To enhance visual comparison, we aggregate the Xenium gene expression data at these aligned locations to match the Visium resolution (Supplementary Figure 2B) (Methods).

Among the 21 genes exhibiting off-target binding to protein-coding genes, 12 were present in the Visium dataset that had at least one corresponding off-target gene also detected in the dataset. For genes with no predicted off-target binding based on perfect sequence homology such as MS4A1, we observed a visually similar spatial pattern between the two technologies (Figure 2A), suggesting that spatially aligned groups of cells across the two technologies express this gene at comparable relative magnitudes. However, for genes with predicted off-target binding based on perfect sequence homology such as TUBB2B, we observed a visually dissimilar spatial pattern between the two technologies (Figure 2B). Importantly, its predicted off-target gene, TUBB2A, in Visium shows a visually more similar spatial pattern to the Xenium TUBB2B. To better visualize the effect of off-target binding within the Xenium data, we aggregated the expression of each gene along with its predicted off-targets found in the Visium dataset and visually compared across spatial locations. Notably, the spatial pattern of the aggregated expression of TUBB2B and TUBB2A in Visium is visually more similar to the spatial pattern of TUBB2B in Xenium, consistent with the prediction that Xenium TUBB2B probes exhibit off-target binding to TUBB2A. We further visually confirmed using the Integrative Genomics Viewer (IGV) that a probe sequence intending to bind to TUBB2B also perfectly aligns to a sequence in TUBB2A, consistent across all annotations evaluated (Supplementary Figure 1B).

Comparison of spatial gene expression patterns between Xenium and Visium.

(A) Spatial gene expression of MS4A1 overlaid on the corresponding histological images for Xenium and Visium. (B) Gene expression patterns for TUBB2B: Xenium expression, Visium expression, the aggregated Visium expression combining TUBB2B and its predicted off-target gene’s expression TUBB2A, and Visium expression of TUBB2B’s predicted off-target TUBB2A. (C) Scatterplot of log-transformed total expression counts (with a pseudocount) for 307 genes comparing Visium and Xenium data. The dotted line represents X = Y, and points (genes) are colored by probe information.

Overall, when comparing the total gene expression between the two technologies, we observed a generally strong positive correlation, consistent with the previously published work (Janesick et al. 2023). We do not observe an obvious trend between gene expression magnitude and the presence of predicted off-target probes (Figure 2C), suggesting that off-target binding prediction alone does not explain the observed higher expression magnitude in Xenium compared to Visium, which may still be attributed to variation in detection efficiency, sequencing depth, and other factors.

Comparison with scRNA-seq reveals single-cell gene expression patterns consistent with off-target binding

To further investigate the potential effects of our predicted off-target binding for Xenium probes, we compared the detected single-cell gene expression patterns in the same previously published work: again, the one assayed with Xenium using the Xenium v1 Human Breast Gene Expression Panel, and another assayed using Chromium Next GEM Single Cell 3’ (Janesick et al. 2023). Briefly, single-cell RNA sequencing (scRNA-seq) with 3ʹ end capture is a technique used to profile gene expression at the single-cell level by profiling the 3ʹ ends of mRNA transcripts with sequencing followed by alignment to a genome or transcriptome for quantification. While this approach provides single-cell resolution gene expression quantification, it lacks spatial information. To enable a single-cell comparison with Xenium, we use Harmony (Korsunsky et al. 2019) to remove batch effects and project cells into a shared UMAP embedding (Supplementary Figure 3) (Methods).

Among the 21 genes exhibiting off-target binding to protein-coding genes, 16 were present in the sc-RNAseq dataset that had at least one corresponding off-target gene also detected in the dataset. Again, for genes with no predicted off-target binding based on perfect sequence homology such as MS4A1, we observed a visually similar gene expression pattern in the harmonized UMAP across both technologies (Figure 3A), suggesting that transcriptionally similar clusters of cells or cell-types across the two technologies express this gene at comparable relative magnitudes. Likewise, again, for genes with predicted off-target binding based on perfect sequence homology such as TUBB2B, we observed a visually dissimilar gene expression pattern on the harmonized UMAP (Figure 3B). Again, its predicted off-target gene, TUBB2A, in scRNA-seq showed a visually more similar expression pattern in the harmonized UMAP embedding to the Xenium TUBB2B. To better illustrate the impact of the off-target probes, we again aggregated the expression of a gene and its predicted off-target genes present in the scRNA-seq data and visually compared across the harmonized UMAP embedding. The aggregated expression of TUBB2B and TUBB2A in the scRNA-seq data shows a visually more similar gene expression pattern in the harmonized UMAP embedding to TUBB2B in the Xenium, consistent with the prediction that Xenium TUBB2B probes exhibit off-target binding with these paralogs (Figure 3B, Supplementary Figure 4).

Comparison of single-cell gene expression patterns between Xenium and scRNA-seq.

(A) Harmonized UMAP visualization of MS4A1 expression for Xenium and scRNA-seq data. (B) Comparison of TUBB2B expression patterns on harmonized UMAP: Xenium expression, scRNA-seq expression, an aggregated scRNA-seq profile combining TUBB2B and its predicted off-target gene’s expression TUBB2A, and scRNA-seq expression of TUBB2B’s potential off-target TUBB2A. Note that TUBB2B also has two additional off-targets present in the scRNA-seq data (Supplementary Figure 4). (C) Scatterplot of log-transformed total expression counts (with a pseudocount) for 313 genes between Visium and scRNA-seq data. The dotted line represents X = Y, and points (genes) are colored by probe information.

Overall, when comparing total gene expression between the two technologies, we again observed a generally strong positive correlation (Figure 3C), similar to the Visium comparison results and consistent with the previously published work (Janesick et al. 2023).

OPT results when allowing mismatches at the terminal ends of the probe sequences identifies additional off-target candidates

Thus far, we have focused on predicting off-target binding based on perfect sequence homology. However, we reason that imperfect sequence matching could still result in off-target binding. Specifically for the Xenium v1 Human Breast Gene Expression Panel, padlock probes with two 20 base pair (bp) arms bind complementary mRNA regions, forming a 40bp probe sequence. A ligase then circularizes the padlock probe, favoring specific 2-bp junctions. Importantly, if there is a sequence mismatch, particularly outside the ligation site towards the terminal ends of the probe sequence, hybridization may still occur and result in off-target binding, albeit with reduced hybridization efficiency (Supplementary Figure 5A). We therefore added an option in OPT to allow imperfect alignments at the ends of the probe sequences, specifying the sequence length at either ends, where mismatches, insertions, deletions, or clipping can occur (Methods).

Allowing for 6bp mismatches on either end of the 40bp probe sequence (i.e., requiring a 28bp match covering the middle of the probe sequence including the ligation site) revealed nine additional genes with potential off-target binding when using GENCODE v47 (Supplementary Table 5). Additionally, 18 of the 45 genes previously predicted to be affected by off-target binding based on perfect sequence homology were now predicted to have additional off-target binding, either through the accumulation of new probes with predicted off-target binding or from the identification of additional predicted off-target genes (Supplementary Table 6). Among these additional genes included CEACAM8, for which we observed a visually dissimilar spatial pattern between Xenium and Visium (Supplementary Figure 6A). Furthermore, the spatial pattern of the aggregate of CEACAM8 with its predicted off-target genes (CEACAM5, CEACAM6, CEACAM7, PSG6) in Visium more closely resembled the spatial pattern of CEACAM8 in Xenium. Likewise, a similar trend is observed in the scRNA-seq comparison (Supplementary Figure 6B). Ultimately, these findings suggest that off-target binding, even with imperfect sequence matching, can contribute to the expression patterns observed in Xenium.

Discussion

Our study presents evidence of off-target probe binding that may distort gene expression profiles affecting at least 21 out of the 280 genes in the Xenium v1 Human Breast Gene Expression Panel. This finding is supported by spatial and single-cell comparative analyses using Xenium with serial section datasets from Visium CytAssist and 3’ single-cell RNA-seq respectively. Although we focus here on the Xenium v1 Human Breast Gene Expression Panel from 10x Genomics, we do not exclude the possibility that off-target binding may similarly affect other probe-based gene detection approaches. To assist in the interpretation of existing probe-based gene expression data as well as future probe design, we provide OPT as a software tool for predicting potential off-target probe binding.

Although we have predicted off-target binding based on sequence alignment, its effect on gene expression quantification may still vary. One reason is that the off-target protein-coding or non-protein-coding gene may not be expressed. For example, although CEACAM6 probes have predicted off-target binding to CEACAM3 based on perfect sequence homology, the sparse expression of CEACAM3 in both the Visium and scRNA-seq breast cancer data led to only a minor difference in the aggregated expression (Supplementary Figure 7). As such, the extent of the effect of off-target binding depends on the expression of the off-target gene and will therefore vary across tissues. On the other hand, self-hybridization or interactions between probes themselves may occur to induce non-specific signals (Supplementary Fig 5B). In general, probe binding specificity is influenced by numerous factors, with many methods previously developed to aid in the design of probe sequences while taking these factors into consideration (Wang and Seed 2003; Rouillard, Zuker, and Gulari 2003; Wernersson and Nielsen 2005; Chou 2010; Li et al. 2011; Hu et al. 2020; Hershberg et al. 2021; Fornace et al. 2022).

Notably, our evaluation was limited to the base 280-gene Xenium v1 Human Breast Gene Expression Panel even though the expression of 313 total genes were quantified in these datasets from Janesick et al. This is because 33 of the 313 targeted genes were probed using custom sequences that were not publicly disclosed. Our comparative analysis suggests that HDC, a gene with custom probe sequences mentioned in the paper, may also be subject to off-target binding. In Xenium, HDC exhibited a distinct spatial pattern and high global expression level, whereas both Visium and scRNA-seq data showed a minimal spatial pattern and sparse expression level respectively (Supplementary Figure 8). We therefore speculate that HDC probes may exhibit off-target binding but cannot fully validate without the probe sequences used for HDC. As such, we emphasize the importance of publishing full probe sequences to enable reproducible science.

We also observed that 34 / 4,809 breast panel probe sequences did not align to any reference transcripts in the GENCODE v47 annotation. 33 of these unmapped probe sequences matched intronic sequences, which would not normally appear in mature transcripts. We manually aligned the one remaining probe sequence (ENSG00000198851|CD3E|02369ef) to the GRCh38 genome and found that it matched to chr11: 118304577 – 118304616, an intergenic region just upstream of CD3E, the target gene for this probe. We suspect that these intron and intergenic panel probe sequences were designed using an earlier version of the GENCODE annotation in which those regions were annotated as exonic. This finding illustrates the importance of disclosing the specific annotation version to promote reproducibility, as well as the ongoing variability of human gene annotation. Likewise, as evidenced by our analysis across GENCODE, RefSeq, and CHESS, we emphasize the variation across these reference annotations and therefore recommend evaluating probe sequences for off-target effects using multiple annotations to be more comprehensive.

Given these challenges, we advise probes with predicted off-target binding to protein-coding genes based on perfect sequence homology be discarded in future experiments. Likewise, we encourage the use of tools like OPT to aid in future probe design decisions and help ensure that probes are optimized to minimize off-target binding based on current transcriptome annotations. For datasets that have already been generated using probes with predicted off-target binding, we generally recommend taking into consideration these predictions to avoid drawing misleading conclusions. For example, we recommend expression measurements for genes with predicted off-target binding be omitted from training foundation models to avoid error propagation. Alternatively, when performing integrative analyses that compare or align gene expression with measurements across technologies, it may be necessary to incorporate off-target binding predictions. For example, integration could be performed between the observed Xenium gene expression and the aggregated expression of the target and predicted off-target genes for the orthogonal technology. Finally, existing literature that base conclusions on genes with predicted off-target binding should be interpreted with caution.

We emphasize that these findings were missed in the previous Janesick et al. publication from 10x Genomics (Janesick et al. 2023). Consistent with previously published observations, we observed a highly correlated total gene expression magnitude between Xenium and Visium as well as scRNA-seq. However, a notable exception is APOBEC3B, which is not expressed according to both Visium and scRNA-seq but highly expressed according to Xenium (Supplementary Figure 9) – a discrepancy that Janesick et al. omitted. We emphasize that positive significant average gene expression correlation is a necessary but not sufficient metric for consistency across technologies and that individual data points should be scrutinized. Likewise, validation with orthogonal technologies could have helped identify discrepancies suggestive of off-target effects. We note Janesick et al. only use immunofluorescence to validate two genes, ERBB2 and MS4A1, which by our analysis were predicted to exhibit no off-target binding. We therefore stress the importance of more thorough orthogonal validations. Although Xenium incorporates blank and negative control probes that are intended to help quantify the rate of non-specific and potential off-target binding, our findings suggest that relying solely on such probes for error detection may be insufficient. Implementing probe redundancy, where the same gene is targeted using different codewords, could provide an additional internal control to enable the detection of off-target binding.

This is not the first instance in which a commercially available platform has encountered challenges in probe design (McCartney et al. 2016; Harbig 2005; Mecham et al. 2004; Liu, Bebu, and Li 2014). These findings underscore the critical role of academic researchers towards ensuring the robustness of industry-led product development by providing oversight, free of financial conflicts of interest through independent federal funding. This complementarity between industry and academia fosters a more rigorous, transparent, and reliable scientific process, ultimately to the benefit of consumers and the public. By shedding light on putative off-target probe binding as well as by providing a tool to enable such off-target binding predictions, this work will help enhance the quality of spatial transcriptomics data and improve the overall reproducibility in spatial transcriptomics research.

Methods

Off-target Probe Tracker Tool (OPT)

OPT (Off-target Probe Tracker) is a Python program that runs nucmer (Marçais and Kingsford 2011) for alignment and then processes the results to predict probe binding based on sequence homology. OPT is available as an open-source Python toolkit at https://github.com/JEFworks-Lab/off-target-probe-tracker. When a user provides a query probe sequence file, a target transcript sequence file, and the annotation used to extract these transcripts, OPT outputs which gene each probe is likely to bind to. Nucmer is a fast nucleotide sequence aligner that uses maximal exact matches as anchors which it then joins together to find longer alignments. By default, OPT saves nucmer results in SAM format and finds perfect sequence matches between a query probe and a target transcript, requiring that alignments consist of only matches and cover the entirety of the query. OPT consists of three modules: (1) flip for reverse complementing probe sequences aligned to the opposite strand of their target genes; (2) track for aligning probe sequences and processing alignment results; and (3) stat for compiling summary statistics on the number of off-target binding probes and affected genes.

In the case that a probe’s target gene has synonyms, we consider alignments to genes annotated with one of its synonym to still be on-target. For example, if a probe that targets NARS shows alignments to a gene called NARS1, we don’t consider it to be off-target binding. We gathered relevant gene synonym relationships using the GeneCards and HGNC online database.

OPT also provides a “pad” mode in which imperfect alignments are allowed at either end of the query (i.e., probe sequence). The -pl parameter sets the pad length at either end of the query and OPT allows for any number of mismatches in these padded regions. For example, if the pad length is 6 and the probe sequence length is 40, then the middle 28bp are the only part of the probe sequence required to match. As long as the critical region is intact, OPT reports an off-target binding site based on this alignment. By default, -pl is set to 0, and the pad mode is activated by providing a non-zero integer to -pl.

Obtaining probe sequences for the Xenium v1 Human Breast Gene Expression Panel

To identify potential off-target binding impacting the 10x Genomics Xenium v1 Human Breast Gene Expression Panel, we obtained the FASTA file of probe sequences from the 10x Genomics Website (https://cdn.10xgenomics.com/raw/upload/v1684948097/software-support/Xenium-panels/hBreast_panel_files/xenium_human_breast_gene_expression_panel_probe_sequences.fasta) and available as Supplementary Data for preservation.

The target gene name and ID was extracted from the probe IDs of the following format:

> gene_id|gene_name|accession

We expected the provided FASTA file to contain probe sequences to be the reverse complemented sequence of their intended target genes and hence align to the reverse strand of their target isoforms. However, when we aligned the breast panel probe sequences to the GENCODE basic (v47) reference transcripts using nucmer, we found that only 2,508 / 4,809 (52.2%) of probe sequences aligned on the reverse strand of their target transcripts (i.e., isoforms of their target genes). For consistency, we enforced that all probe sequences be oriented in the same direction and align to the forward strand of their target genes and transcripts. As such, we reverse-complemented these 2,508 probe sequences. We also added this functionality as an OPT module called “flip” in which probe sequences aligned to the reverse strand of their targets are reverse complemented. We expect probe sequences to align to the forward strand of transcripts (i.e., both oriented in the same direction) during the downstream probe sequence binding prediction step.

Visium Comparison

The Visium CytAssist dataset, collected from a breast cancer tissue block utilized in Janesick et al., was downloaded from the 10x Genomics website (https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast). This dataset originally contained 4,992 spots with x-y coordinates and includes 18,085 genes per spot. Six genes (AKR1C1, ANGPT2, BTNL9, CD8B, POLR2J3, TPSAB1) were excluded from the analysis since they were not present in the Visium dataset.

To compare spatial gene expression patterns from Visium and Xenium technologies, we first mapped all the data to the same coordinate space. We used STalign (v1.0.1), a computational tool that utilizes affine transformations along with diffeomorphic metric mapping to align target and source datasets (Clifton et al. 2023). The initial alignment involved only affine transformations and eight manually determined landmarks to align the Visium histology image (source) to the Xenium histology image (target). This transformation brought the Visium image into the coordinate space of the higher-resolution Xenium image. We then applied this learned transformation to the Visium spots, ensuring that they were correctly positioned relative to both histology images. Next, we used STalign to map the Xenium transcripts (source) onto their corresponding Xenium histology image (target) using both affine and diffeomorphic metric mapping. The transcripts were rasterized at 30μm resolution, with an initial affine transformation guided by four manually defined landmarks. Diffeomorphic metric mapping was then performed with the following parameters: a = 2500, epV = 1, niter = 2000, sigmaA = 0.11, sigmaB = 0.10, sigmaM = 0.15, sigmaP = 50, muA = [1, 1, 1], muB = [0, 0, 0], with all other settings left at their defaults. We extracted the overlapping regions between the two datasets (Supplementary Figure 2A), which reduced the total spots in the Visium dataset to 3,958. Finally, we aggregated the Xenium gene expression data to ∼55μm x 55μm patches that correspond to the spatial locations of the Visium spots, resulting in matched-resolution spatial gene expression for both technologies (Supplementary Figure 2B).

Single-cell RNA-seq Comparison

The Chromium Next GEM 3’ scRNA-seq dataset, collected from a breast cancer tissue block utilized in Janesick et al., was downloaded from the 10x Genomics website (https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast). This dataset contained 12,388 cells with 36,601 genes per cell. All 313 unique genes present in the Xenium dataset are also in the scRNA-seq data, hence both datasets were subsetted to these genes for this analysis.

Both scRNA-seq and Xenium provide single-cell resolution data. To integrate these datasets, we first removed cells lacking detectable gene expression. We the normalized the combined gene expression data using counts per million (CPM) and applied a log transformation with a pseudocount of 1. Principal component analysis (PCA) is then applied to the normalized data, and batch effects are corrected using Harmony (v1.2.3) on the top 30 principal components (PCs) using default parameters except for theta, which was set to 8, to promote further mixing with clusters across technologies. Finally, Uniform Manifold Approximation and Projection (UMAP) is performed on the harmonized PCs, generating a shared 2D embedding across the two technologies, and the data is further facetted by technology for visualization (Supplementary Figure 3).

Cross-annotation analysis

To compare OPT’s results with different reference annotations, we used the most recent releases of GENCODE basic (v47), GENCODE comprehensive (v47), RefSeq (v110), and CHESS (v3.1.3) annotation of the GRCh38 genome. Note that GENCODE “basic” is the more reliable version of the annotation and is much closer to RefSeq and CHESS. GENCODE “comprehensive” includes hundreds of thousands of low-quality annotations, which we included in some of our analyses for completeness. Note also that GRCh38 has many non-reference sequences called “alternative scaffolds”; we removed these for our analysis. We then used gffread to extract transcripts as defined in these annotations by running:

$ gffread -w transcripts.fa -g grch38.p12/14.fa annotation.gff

The GRCh38.p14 assembly was used during transcript sequence extraction for all reference annotations, except for CHESS which specifies that the annotation maps genes and transcripts onto the GRCh38.p12 assembly. For RefSeq, we renamed the VD(J) segment features as transcript features to ensure consistency, and we also removed transcript sequences with the gene_biotype “pseudogene.” RefSeq has a separate biotype called “transcribed_pseudogene,” but doesn’t annotate transcripts for these features. We considered transcripts annotated for a small subset of just pseudogenes an error in the annotation.

eLife Assessment

This valuable study identifies and characterizes probe binding errors in a widely used commercial platform for visualizing gene activity in tissue samples, discovering that at least 21 out of 280 genes in a human breast cancer panel are not accurately detected. The authors provide convincing evidence for their findings validated against multiple independent sequencing technologies and reference datasets. Given the broad adoption of this spatial gene detection platform in biomedical research, this work provides an essential quality control resource that will improve data interpretation across numerous studies.

Reviewer #1 (Public review):

Summary:

In the manuscript, Hallinan et al. describe off-target probe binding in the 10x Genomics Xenium platform, which results in invalid profiling of some genes in a spatial context. This was validated by comparing the Xenium results with Visium and scRNA-seq using human breast tissue, which are comprehensive and convincing. The authors also provide a dedicated tool to predict such off-target binding, Off-target Probe Tracker (OPT), which could be widely adopted in the field by researchers who are interested in validating the previously published results.

Strengths:

(1) This is the first study to suggest off-target binding of probes in the gene panels of the Xenium platform, which could be easily overlooked.

(2) The results were rigorously validated with two different methods.

(3) This paper will be a helpful resource for properly interpreting the results of previously published papers based on the Xenium platform (the signals could be mixed).

Weaknesses:

(1) The results were only tested with one tissue (human breast). However, this is not a major weakness, as one can easily extrapolate that this should be the case for any other tissue.

(2) Once the 10X Genomics corrects their gene panels according to this finding, the tool (OPT) will not be useful for most people. Still, it can be used by those who want to design de novo probes from scratch.

Reviewer #2 (Public review):

This paper describes an analysis of a commercially available panel for a spatial transcriptomic approach and introduces a computational tool to predict potential off-target binding sites for the type of probe used in the aforementioned panel. The performance of the prediction tool was validated by examining a dataset that profiled the same cancer tissue with multiple modalities. Finally, a detailed analysis of the potential pitfalls in a published study communicated by the company that commercialized the spatial transcriptomic platform in question is provided, along with best practice guidelines for future studies to follow.

Strengths:

The manuscript is clearly written and easy to follow.

The authors provide clean, organized, and well-documented code in the associated GitHub repository.

Weaknesses:

The manuscript section on the software tool feels underdeveloped.

Reviewer #3 (Public review):

Summary:

The authors present a new computational method (OPT) for predicting off-target probe binding in the commercial 10X Xenium spatial transcriptomics platform. They identified 28 genes in the 10x xenium human breast cancer gene panel (280 genes) that are not accurately detected at the single-molecule level. They validated the predicted off-target binding using reference data from single-cell RNA-seq and 3'-sequencing-based Visium RNA-seq. This work provides a practical resource and will serve as a valuable reference for future data interpretation.

Strengths:

(1) Provides a toolbox for the community to identify off-target probes.

(2) Validates the predictions using single-cell RNA-seq and sequencing-based Visium RNA-seq datasets.

Weaknesses:

(1) Does not apply the OPT method to the most widely used Xenium gene panels (e.g., pan-Human, pan-Mouse panels with ~5,000 genes each).

(2) Lacks clarity on how the confidence level of off-target predictions is calculated.

Author Response:

The following is the authors’ response to the original reviews

Public Reviews:

We thank the editors and the reviewers for their constructive feedback in helping us strengthen this manuscript.

During the revision process, new information was shared with us by the 10x Genomics team regarding the Xenium probe sequences evaluated in our original paper. Briefly, the Xenium probe sequences we evaluated represented an earlier iteration of the probes used to generate the data in Janesick et al. Further, we were made aware that the probe sequences used in Janesick et al. represented an earlier iteration of the commercially available Xenium v1 Human Breast Gene Expression Panel. We now elaborate further in a new Supplementary Note. We have therefore updated the paper throughout to reflect this new understanding, though we emphasize that our conclusions do not change. Rather, this newfound understanding provides stronger evidence of off-target probe binding with imperfect sequence matching, which we support with new supplementary figures.

(1) Limited evaluation of tissues and gene panels

“The results were only tested with one tissue (human breast). However, this is not a major weakness, as one can easily extrapolate that this should be the case for any other tissue.”

“Does not apply the OPT method to the most widely used Xenium gene panels (e.g., pan-Human, pan-Mouse panels with ~5,000 genes each).”

“The authors claim that OPT is a generalizable method for identifying off-target probes. To support this claim, they should provide similar predictions for the Xenium Pan-Human or Pan-Mouse gene panels, which are more widely used than the breast cancer panel.”

“While I understand that conducting new experimental studies is likely beyond the authors' intended scope of the manuscript, the narrow reliance on Janesick et al. for all of the validation makes it difficult to assess the broad usability of OPT. In the absence of designing and then validating novel padlock probe designs with OPT, are there other publicly available datasets that authors could perform secondary analysis on using OPT?”

Our primary focus on breast cancer was driven by data availability rather than tissue specificity. For this probe panel, matched Xenium, Visium, and scRNA-seq datasets are publicly available, enabling direct cross-platform comparisons of gene expression and allowing us to evaluate the impact of off-target probe binding in Xenium.

OPT is tissue-agnostic and can be applied to any probe panel regardless of tissue type. To demonstrate this generalizability, we have now applied OPT on all publicly available 10x Genomics probe sets beyond the breast panel, including the Xenium pan-Human and pan-Mouse gene panels. The complete results of these analyses have been generated and are provided as a compressed zip file accompanying the revised manuscript.

Beyond pre-designed panels, in this revision, we have now also applied OPT to custom Xenium gene panels from the Human BioMolecular Atlas Program (HUBMAP) and further demonstrate integration of HUBMAP RNA-seq data to evaluate the impact of potential predicted off-targets in a new section “Bulk RNA-seq reference atlases suggest off-target binding can variably impact results in Xenium custom probe panels.”

Overall, in these newly evaluated panels, we identify many cases of off-target probe binding with non-negligible expression of off-target genes in the target tissue, underscoring that our findings are not specific to human breast tissue. Therefore, in the revision, we have broadened the title to “Evidence of off-target probe binding affecting 10x Genomics Xenium Gene Panels compromise accuracy of spatial transcriptomic profiling”

(2) Limited quantifications

“Lacks clarity on how the confidence level of off-target predictions is calculated.”

“How can the confidence level of these off-target predictions be quantitatively assessed? Please provide benchmarks or validation metrics if available.”

We thank the reviewer for raising this important point. To strengthen our claim that predicted off-targets can contribute to observed Xenium expression patterns, we incorporated a quantitative assessment in addition to the qualitative comparisons presented previously. Specifically, we leveraged Visium and scRNA-seq data to compare spot- and cluster-level expression of target genes alone versus expression aggregated with their predicted off-target genes. Across all examples shown, inclusion of predicted off-targets consistently resulted in stronger agreement with the Xenium results, as reflected by decreased RMSE and increased Pearson correlation relative to using the target gene alone.

We emphasize, however, that OPT does not assign a formal confidence score to off-target predictions based on sequencing data alone. Importantly, identification of a potential off-target by OPT does not imply that it will necessarily affect Xenium results. As we’ve noted, if the off-target gene is not expressed, then it will not affect the observed gene expression magnitudes of the target gene. To help users assess whether predicted off-target genes will affect observed gene expression magnitudes of the target gene for a tissue of interest, we now provide a complementary analysis, including heat-map visualizations comparing the expression of target genes and their predicted off-targets in matched bulk RNA-seq or scRNA-seq datasets from the same tissue (Supplementary Figures 9, 10, 11). We hope this evaluation pipeline will clarify to researchers they can evaluate whether predicted off-targets will appreciably affect results in their tissue of interest.

(3) Under-developed and non-essential software

“The manuscript section on the software tool feels underdeveloped.”

“Once the 10X Genomics corrects their gene panels according to this finding, the tool (OPT) will not be useful for most people. Still, it can be used by those who want to design de novo probes from scratch.”

“Since the authors claim that OPT is intended for community use, the paper should provide a clear, step-by-step user guide, such as Jupyter tutorial, ideally as supplementary material.”

We agree with the reviewers that the description of the software tool itself is relatively concise. This is intentional, as the primary goal of this manuscript is not to introduce a standalone software framework, but rather to use the tool as a means to characterize and quantify off-target probe binding and its potential downstream impact on spatial gene expression analyses. Accordingly, our emphasis is placed on the biological and analytical insights enabled by this approach, rather than on extensive software tool details. To support potential users, we have now included additional software documented with an example Python notebook demonstrating how it can be applied to any probe panels in the GitHub repository: https://github.com/JEFworks-Lab/off-target-probe-tracker/blob/main/example.ipynb

Likewise, the primary goal of this manuscript is not to suggest that a specific vendor’s probe panels are flawed, but rather to demonstrate that off-target probe binding is a general and underappreciated phenomenon that can occur in some probe-based spatial transcriptomics platforms to meaningfully impact downstream analyses and biological interpretation.

OPT was developed as a framework to identify potential off-target probe interactions based on sequence homology. In practice, OPT can serve as a post hoc tool that allows researchers to assess whether predicted off-target interactions may exist in a given panel and to account for these possibilities when interpreting spatial expression patterns, even when panels have been developed by the many probe designing methods now highlighted in the revised manuscript. Given the complexity of probe design and hybridization behavior, we believe that explicitly identifying and reporting potential off-targets remains valuable for downstream data interpretation, cross platform comparisons, and reproducibility. Thus, OPT is intended to complement existing probe design strategies and vendor efforts, rather than replace them, by providing researchers with additional context to interpret their data more accurately.

In our revision, we have therefore elaborated on this in the discussion, reiterated here for convenience: “Although we focus here on the 10x Genomics Xenium technology, we do not exclude the possibility that off-target binding may similarly affect other probe-based gene detection approaches from other commercial vendors. Any technology that relies on hybridization-based detection is inherently susceptible to off-target probe binding when sequence similarity exists. Further, hybridization-based detection often inherently involves a trade-off between sensitivity and specificity. Given these inherent technological limitations, we therefore emphasize the importance of transparency through sharing probe sequences. However, many companies do not release the probe sequences used in their assays, limiting the consumer’s ability to fully interpret their results as well as the community’s ability to effectively characterize and benchmark performance variation across platforms. Therefore, we strongly recommend that companies publish probe sequences for pre-designed panels and likewise that researchers using these technologies should obtain and publish probe sequences used in their studies to support transparent and reproducible science. “

Recommendations for the authors:

“The paper only describes evidence of the off-target effect based on perfect sequence homology, although the tool (OPT) provides an option to find additional "potential" off-targets that allow mismatches. It would be very nice if the authors could additionally provide at least one example of off-target binding with at least one mismatch.”

We thank the reviewer for the opportunity to clarify this point. In addition to analyses based on perfect sequence homology, we examined predicted off-target binding when allowing mismatches at the terminal ends of probe sequences. This analysis is presented in the Results section titled “OPT results when allowing mismatches at the terminal ends of the probe sequences identifies additional off-target candidates.”

In this revision, we now allowed a 10bp padding on either end of the 40bp probe sequence, permitting imperfect sequence matching at the terminal regions. Under these conditions, OPT identified additional off-target candidates, including TUBB2B and ACTG2, which we highlight as representative examples (Supplementary 7,8). We further demonstrate how these predicted off-target interactions impact gene expression concordance by comparing Xenium measurements with both Visium and scRNA-seq data, showing measurable changes in cross-platform agreement. Together, these results illustrate that allowing mismatches reveals biologically relevant off-target effects beyond those captured by perfect sequence homology alone.

“Clarifications and updates for Figure 2A-B

Xenium offers a resolution of up to 200 nanometers with continuous readout, without pixel gaps. However, the figures shown in Figure 2A-B appear pixelated - why is this the case? Could the authors clarify this discrepancy and, if possible, provide the raw feature intensity data for Xenium in the supplementary materials?

Additionally, there appear to be no visible gaps in the Visium graphs. Could the authors update the figure panels to represent the true spot locations for Visium, to more accurately reflect the underlying data structure?”

We thank the reviewer for the opportunity to clarify these points. The goal of Figure 2A-B is to facilitate a direct visual comparison of gene expression patterns between the Visium and Xenium platforms. To enable this comparison, we aggregated the single-cell Xenium data into spatial patches matching the effective resolution of Visium spots (55x55µm). Similarly, Visium spots were rendered as patches to produce a more continuous visual representation. As a result of this aggregation and visualization choice, the Xenium expression plots appear pixelated despite Xenium’s native subcellular resolution (up to ~200 nm with continuous readout). We have clarified this processing and visualization step in the Methods to avoid confusion.

With respect to the Visium expression plots, the lack of gaps is also a consequence of rendering each spot as a filled patch rather than plotting traditional Visium spots. This was done intentionally to maintain visual consistency with the aggregated Xenium data and to emphasize spatial concordance rather than the underlying sampling geometry. We have now explicitly stated this design choice to improve clarity.

“I found the format of the manuscript to be at times confusing and perhaps a bit of an odd fit for a general interest journal. A significant portion of the manuscript is spent critiquing a specific publication, "High resolution mapping of the tumor microenvironment using integrated single-cell, spatial and in situ analysis" published by Janesick et al. (of 10x Genomics, Inc.) in Nature Communications in 2023. This content would seem more appropriate as a Comment submitted to Nature Communications, potentially to be accompanied by a response from the authors of Janesick et al. at 10x.”

I would like to address this important point as the corresponding author who takes primary responsibility for the unconventional decision to submit this manuscript to eLife as opposed to as a commentary suggested by the reviewer.

Consistent with the reviewer, I did initially consider submitting this as a Matters Arising to Nature Communications. However, after consultation with other senior colleagues and co-authors, I decided to forgo this route on the basis that the information provided in a Matters Arising must be kept confidential. I was concerned that this would lead to long, drawn-out private exchanges. As we note in the manuscript, the Xenium platform's widespread use and high cost imposed a certain urgency that I believed warranted open and rapid dissemination.

Therefore, we submitted to eLife with the hope that eLife’s unique continuous post-publication public peer review process will enable the rapid dissemination of these important financially-sensitive insights while permitting constructive criticisms from both industry and academic expert reviewers to be openly considered by all readers.

Version 2

Abstract

The accuracy of spatial gene expression profiles generated by probe-based in situ spatially-resolved transcriptomic technologies depends on the specificity with which probes bind to their intended target gene. Off-target binding, defined as a probe binding to something other than the target gene, can distort a gene’s true expression profile, making probe specificity essential for reliable transcriptomics. Here, we investigated off-target binding affecting the 10x Genomics Xenium technology. We developed a software tool, Off-target Probe Tracker (OPT), to identify putative off-target binding via alignment of probe sequences and assessing whether mapped loci corresponded to the intended target gene across multiple reference annotations. Applying OPT to a Xenium human breast gene panel, we identified at least 14 out of the 313 genes in the panel potentially impacted by off-target binding to protein-coding genes. To substantiate our predictions, we leveraged a Xenium breast cancer dataset generated using this gene panel and compared results to orthogonal spatial and single-cell transcriptomic profiles from Visium CytAssist and 3ʹ single-cell RNA-seq derived from the same tumor block. Our findings indicate that for some genes, the expression patterns detected by Xenium demonstrably reflect the aggregate expression of the target and predicted off-target genes based on Visium and single-cell RNA-seq rather than the target gene alone. We further applied OPT to identify potential off-target binding in custom gene panels and integrate tissue-specific RNA-seq data to assess effects. Overall, this work enhances the biological interpretability of spatial transcriptomics data and improves reproducibility in spatial transcriptomics research.

Introduction

Recent advances in high-throughput spatially-resolved transcriptomic profiling technologies have enabled the investigation of gene expression with high spatial resolution within tissues. One such commercially available spatial transcriptomics platform is Xenium from 10x Genomics, a publicly traded company with a market capitalization exceeding $2 billion as of November 2025 (Yahoo Finance 2025). Xenium achieves spatial gene expression profiling at single-cell resolution for targeted genes using a probe-based in situ detection approach. 10x Genomics currently offers targeted gene panels with pre-designed probe sets. As of December 2024, over 16,000 Xenium consumable reactions have been sold, with each tissue slide profiled costing approximately $5,000, underscoring the platform’s widespread use and high commercial value (10x Genomics 2025b; 2025a).

Briefly, Xenium uses padlock probes that include sequences complementary to the RNA of target genes. Once a padlock probe binds to its target, it is ligated and subsequently amplified via rolling circle amplification (RCA). Fluorescently labeled decoder probes then hybridize to the amplified RCA product, enabling the simultaneous detection and decoding of the optical signature, or codeword, specific to each gene in the panel through successive rounds of fluorescence imaging. When combined with cell-segmentation, this approach allows for spatially-resolved single-cell quantification of gene expression.

The accuracy of these gene expression measurements thus relies on the specificity of the probes to bind to their intended target gene. We define off-target binding as when a probe binds to something other than the RNA sequence intended to correspond to the target gene (Figure 1). We note once ligation and RCA occur, the resulting fluorescent signal cannot be distinguished between on-target and off-target binding. As such, off-target binding can distort the quantification of the intended target gene’s expression, as the observed expression would represent a combination of the target as well as off-target expression.

Schematic of potential off-target binding in 10x Genomics Xenium.

In this illustration, the arms of the padlock probes were designed to bind an RNA sequence intended to correspond to a target gene (green). However, these probes exhibit off-target binding and bind to an RNA sequence in a different off-target gene (red). The probe is circularized and subsequently amplified via rolling circle amplification (RCA). Hybridization of fluorescent probes to the RCA product enables the generation a fluorescent signal that is used to quantify RNA expression within cells.

To predict for such potential off-target binding, we developed Off-target Probe Tracker (OPT), a software tool that aligns probe sequences to an annotated transcriptome with the option to allow for mismatches that may still permit probe binding. Using OPT, we identify putative off-target probe binding to protein-coding genes affecting at least 14 out of 313 genes in a 10x Genomics Xenium human breast gene panel, compromising the accuracy of their spatial transcriptomic profiles. We substantiate our predictions using data from orthogonal spatial and single-cell gene expression profiling technologies. We further apply OPT to identify potential off-target binding in custom gene panels and integrate tissue-specific RNA-seq data from The Human BioMolecular Atlas Program (HuBMAP) to assess whether such off-target binding could meaningfully affect assayed expression patterns in specific tissues. By facilitating a more rigorous evaluation of probe sequence specificity, tools like OPT can aid in future probe design decisions to help ensure that probes are optimized to minimize off-target binding based on current transcriptome annotations.

Results

OPT predicts potential off-target probe binding

To identify potential off-target binding impacting the 10x Genomics Xenium technology, we require the probe sequences for a specific gene panel of interest, generally represented in a FASTA file. To this end, we first focus on a human breast gene panel used in the Janesick et al. publication, courtesy of 10x Genomics (Methods; Supplementary Table 1). This file includes 2,582 probe sequences that are 40bp in length and designed to target 313 genes, including 33 genes targeted by custom probes, with an average of 8 probe sequences per gene (ranging from 2 to 21 probe sequences per gene). We note this panel represents an earlier iteration of and is highly similar to the commercially available pre-designed Xenium v1 Human Breast Gene Expression Panel (Supplementary Note).

To enable the prediction of potential off-target binding, we developed a software tool called OPT (Methods) that uses nucmer (Marçais et al. 2018) to align probe sequences to various reference transcriptomes, which comprise curated collections of transcript isoforms for all genes in a species. OPT features adjustable parameters for binding strictness (e.g., number of mismatches) and generates a summary file that details all targeted genes along with their potential off-targets based on the sequence alignments. 10x Genomics designed its probe sequences using the GENCODE “basic” annotation (Mudge et al. 2025), so initially we also used the latest GENCODE “basic” annotation (v47) to predict off-target binding for these probes.

We first sought to predict if a probe has off-target binding based on perfect sequence homology (i.e., if it aligns with 100% identity) with any annotated transcripts other than those that belong to the intended target gene. Of the 2,582 probe sequences in this gene panel, using GENCODE v47, OPT identified 121 probe sequences across 37 genes as having off-target binding based on perfect sequence homology (Table 1). Among the 37 genes with predicted off-target binding, the number of affected probe sequences per gene ranged from 1 to 8. Overall, these off-target probes matched 71 other genes, including 20 protein-coding genes, 31 pseudogenes, 10 long non-coding RNA, 9 transcripts labeled as nonsense-mediated decay, and one microRNA gene.

Off-target binding predictions vary across different annotations

OPT relies on alignments to an annotated transcriptome, which ideally reflects all genes and gene variants stably transcribed in a given species. However, genome annotation is still an active area of research (Varabyou et al. 2023), with discrepancies across annotations in gene counts, isoforms, and many other features. We therefore further used OPT to predict for off-target binding and affected genes using two additional human genome annotation sets, RefSeq (v110), (O’Leary et al. 2016) and CHESS (v3.1.3) (Varabyou et al. 2023), and compared to our previous results from GENCODE (v47) (Mudge et al. 2025) (Methods).

When considering only perfect sequence homology, while we previously found 37 affected genes using GENCODE, we found 14 when using RefSeq and 23 when using CHESS (Supplementary Table 2, 3). Given that RefSeq and CHESS have more transcripts than GENCODE, these discrepancies in off-target binding predictions was not simply an artifact of the difference in transcript set sizes.

While the human annotation databases mostly agree on the number of protein-coding genes in the genome, they remain widely divergent on pseudogenes and lncRNA genes. Therefore, we focused on how the results change when we restrict our analysis to only protein-coding genes. By excluding pseudogenes (which are presumably not expressed), lncRNAs, and other non-protein-coding RNAs when using OPT, the number of affected genes fell to 11 for GENCODE, 10 for RefSeq and 9 for CHESS (Supplementary Table 4). Again, these discrepancies reflect annotation differences. For example, the probe sequence (ENSG00000196154|S100A4|ab4e3dc), which was designed to target S100A4 based on GENCODE annotations, also aligns to S100A5 in RefSeq (Supplementary Figure 1). We reason that if a probe sequence aligns off-target to a protein-coding gene based on any of these annotations, it could result in off-target binding. We therefore focused further analysis on the union of genes with predicted off-target binding to protein-coding genes across the 3 annotations, resulting in 14 genes: ADH1B, AKR1C1, APOBEC3A, APOBEC3B, AQP1, C1QA, CD79B, CD8B, CEACAM6, POLR2J3, S100A4, TOMM7, TPD52, and TPSAB1 (Supplementary Figure 2; Supplementary Table 5).

Comparison with Visium CytAssist reveals spatial gene expression patterns consistent with off-target binding

To investigate the potential effects of our predicted off-target binding for this Xenium human breast gene panel in experimental settings, we compared spatial gene expression patterns detected in two previously published spatial transcriptomics datasets from serial sections of the same breast cancer tissue: one section assayed with Xenium using this gene panel, and another assayed using Visium CytAssist, an orthogonal spatial transcriptomics platform (Janesick et al. 2023). Briefly, Visium CytAssist is a sequencing-based spatial transcriptomics technology in which RNA is hybridized to spatially barcoded capture spots on a slide, enabling spatial transcriptomic mapping after sequencing. However, while Xenium offers single-cell resolution gene expression quantification, Visium quantifies gene expression within 55μm spots. To enable direct comparison, we first structurally aligned the Xenium and Visium tissue sections using STalign (Clifton et al. 2023), restricting our analysis to overlapping regions since different parts of the tissue were profiled (Supplementary Figure 3A). To improve visual comparability, we aggregated the Xenium gene expression data at the aligned locations to match the Visium spatial resolution and visualized using resolution-matched tiles (Supplementary Figure 3B; Methods).

Among the 14 genes exhibiting off-target binding to protein-coding genes, 4 were present in the Visium dataset that had at least one corresponding off-target gene also detected in the dataset. For genes with no predicted off-target binding based on perfect sequence homology such as MS4A1, we observed a visually similar spatial pattern between the two technologies (Figure 2A), suggesting that spatially aligned groups of cells across the two technologies express this gene at comparable relative magnitudes. This can be quantitively assessed by comparing the pseudo-log gene expression values from both Visium and Xenium at matched spatial locations and computing the RMSE and Pearson correlation in a manner similar to STcompare (Clifton et al. 2025). For MS4A1, the RMSE is relatively low at 3.746, and the Pearson correlation of 0.382 indicates a moderate degree of concordance between the two technologies. However, for genes with predicted off-target binding based on perfect sequence homology such as APOBEC3B, we observed a visually dissimilar spatial pattern between the two technologies (Figure 2B). Consistent with this, the RMSE is relatively high at 5.452, and the Pearson correlation is nan because APOBEC3B is not expressed in the Visium dataset. Importantly, its predicted off-target genes, APOBEC3D and APOBEC3F, in Visium shows a visually more similar spatial pattern to the Xenium APOBEC3B. To better visualize the effect of off-target binding within the Xenium data, we aggregated the expression of each gene along with its predicted off-targets found in the Visium dataset and visually compared across spatial locations. Notably, the spatial pattern of the aggregated expression of APOBEC3B, APOBEC3D and APOBEC3F in Visium is visually more similar to the spatial pattern of APOBEC3B in Xenium. Quantitatively, comparing this aggregated expression results in a decrease in RMSE to 4.465 and a non-nan Pearson correlation of 0.160, consistent with the prediction that Xenium APOBEC3B probes exhibit off-target binding to APOBEC3D and APOBEC3F. We further visually confirmed using the Integrative Genomics Viewer (IGV) where two probe sequences intending to bind to APOBEC3B were found to perfectly align to sequences in both APOBEC3D and APOBEC3F, consistent across all annotations evaluated (Supplementary Figure 4).

Comparison of spatial gene expression patterns between Xenium and Visium.

(A) Spatial gene expression of MS4A1 overlaid on the corresponding histological images for Xenium and Visium, accompanied by a density plot comparing Xenium vs. Visium MS4A1 expression. The dotted line indicates the identity line (X = Y), and the solid line represents the line of best fit. (B) Gene expression patterns for APOBEC3B: Xenium expression, Visium expression, the aggregated Visium expression combining APOBEC3B and its predicted off-target gene’s expression APOBEC3D and APOBEC3F, and Visium expression of APOBEC3B’s predicted off-targets APOBEC3D and APOBEC3F. Two density plots are shown: one comparing Xenium vs. Visium for APOBEC3B alone, and one comparing Xenium vs. the aggregated Visium expression of APOBEC3B with all off-targets. The dotted line indicates the identity line (X = Y), and the solid line represents the line of best fit. (C) Scatterplot of log-transformed total expression counts (with a pseudocount) for 307 genes comparing Visium and Xenium data. The dotted line indicates the identity line (X = Y), and points (genes) are colored by probe information.

Overall, when comparing the total gene expression between the two technologies, we observed a generally strong positive correlation, consistent with the previously published work (Janesick et al. 2023). We do not observe an obvious trend between gene expression magnitude and the presence of predicted off-target probes (Figure 2C), suggesting that off-target binding prediction alone does not explain the observed higher expression magnitude in Xenium compared to Visium, which may still be attributed to variation in detection efficiency, sequencing depth, and other factors.

Comparison with scRNA-seq reveals single-cell gene expression patterns consistent with off-target binding

To further investigate the potential effects of our predicted off-target binding for this Xenium human breast gene panel, we compared the detected single-cell gene expression patterns in the same previously published work using Chromium Next GEM Single Cell 3’ (Janesick et al. 2023). Briefly, single-cell RNA sequencing (scRNA-seq) with 3ʹ end capture is a technique used to profile gene expression at the single-cell level by profiling the 3ʹ ends of mRNA transcripts with sequencing followed by alignment to a genome or transcriptome for quantification. While this approach provides single-cell resolution gene expression quantification, it lacks spatial information. To enable a single-cell comparison with Xenium, we use Harmony (Korsunsky et al. 2019) to remove batch effects and project cells into a shared UMAP embedding (Supplementary Figure 5A-B). We also performed Leiden clustering on the harmonized principal components to quantitively compare cluster expression (Supplementary Figure 5C; Methods).

Among the 14 genes exhibiting off-target binding to protein-coding genes, 10 were present in the scRNA-seq dataset that had at least one corresponding off-target gene also detected in the dataset. Again, for genes with no predicted off-target binding based on perfect sequence homology such as MS4A1, we observed a visually similar gene expression pattern in the harmonized UMAP across both technologies (Figure 3A), suggesting that transcriptionally similar clusters of cells or cell-types across the two technologies express this gene at comparable relative magnitudes. We further quantitatively assessed the data by comparing the pseudo-log gene expression values obtained from the clusters within the clustered harmonized UMAP and computed the RMSE and Pearson correlation. For MS4A1, the RMSE is relatively low at 0.479, and the Pearson correlation of 0.991 indicates a strong degree of concordance between the two technologies. Likewise, again, for genes with predicted off-target binding based on perfect sequence homology such as APOBEC3B, we observed a visually dissimilar gene expression pattern on the harmonized UMAP (Figure 3B). Consistent with this, the RMSE is relatively high at 0.829, and the Pearson correlation is nan because APOBEC3B is not expressed in the scRNA-seq dataset. Again, its predicted off-target genes, APOBEC3D and APOBEC3F, in scRNA-seq showed a visually more similar expression pattern in the harmonized UMAP embedding to the Xenium APOBEC3B. To better illustrate the impact of the off-target probes, we again aggregated the expression of a gene and its predicted off-target genes present in the scRNA-seq data and visually compared across the harmonized UMAP embedding. The aggregated expression of APOBEC3B, APOBEC3D and APOBEC3F in the scRNA-seq data shows a visually more similar gene expression pattern in the harmonized UMAP embedding to APOBEC3B in the Xenium data (Figure 3B). Quantitatively, comparing this aggregated expression results in a decrease in RMSE to 0.596 and a non-nan in Pearson correlation to 0.417, consistent with the prediction that Xenium APOBEC3B probes exhibit off-target binding with these paralogs.

Comparison of single-cell gene expression patterns between Xenium and scRNA-seq.

(A) Harmonized UMAP visualization of MS4A1 expression for Xenium and scRNA-seq data, accompanied by a scatterplot comparing Xenium vs. scRNA-seq MS4A1 cluster expression. The dotted line indicates the identity line (X = Y), and the solid line represents the line of best fit. (B) Comparison of APOBEC3B expression patterns on harmonized UMAP: Xenium expression, scRNA-seq expression, an aggregated scRNA-seq profile combining APOBEC3B and its predicted off-target gene’s expression APOBEC3D and APOBEC3F, and scRNA-seq expression of APOBEC3B’s predicted off-targets APOBEC3D and APOBEC3F. Two scatterplots are shown: one comparing Xenium vs. scRNA-seq for APOBEC3B cluster expression alone, and one comparing Xenium vs. the aggregated scRNA-seq cluster expression of APOBEC3B and it’s predicted off-targets. The dotted line indicates the identity line (X = Y), and the solid line represents the line of best fit. (C) Scatterplot of log-transformed total expression counts (with a pseudocount) for 313 genes between Visium and scRNA-seq data. The dotted line indicates the identity line (X = Y), and points (genes) are colored by probe information.

Overall, when comparing total gene expression between the two technologies, we again observed a generally strong positive correlation (Figure 3C), similar to the Visium comparison results and consistent with the previously published work (Janesick et al. 2023).

OPT results when allowing mismatches at the terminal ends of the probe sequences identifies additional off-target candidates

Thus far, we have focused on predicting off-target binding based on perfect sequence homology. However, we reason that imperfect sequence matching could still result in off-target binding. Specifically, for Xenium v1, which underlies the Xenium human breast gene panel, padlock probes with two 20 base pair (bp) arms bind complementary mRNA regions, forming a 40bp probe sequence. A ligase then circularizes the padlock probe, favoring specific 2-bp junctions. Importantly, if there is a sequence mismatch, particularly outside the ligation site towards the terminal ends of the probe sequence, hybridization may still occur and result in off-target binding, albeit with reduced hybridization efficiency (Supplementary Figure 6A). We therefore added an option in OPT to allow imperfect alignments at the ends of the probe sequences, specifying the sequence length at either ends, where mismatches, insertions, deletions, or clipping can occur (Methods).

Allowing for 10bp mismatches on either end of the 40bp probe sequence (i.e., requiring a 20bp match covering the middle of the probe sequence including the ligation site) revealed 16 additional genes with potential off-target binding when using GENCODE v47 (Supplementary Table 6). Additionally, 9 of the 37 genes previously predicted to be affected by off-target binding based on perfect sequence homology were now predicted to have additional off-target genes (Supplementary Table 7). Among these additional genes were TUBB2B and ACTG2, both of which showed visually dissimilar spatial patterns between Xenium and Visium, accompanied by comparably high RMSE and low Pearson correlation (Supplementary Figure 7A-B). In contrast, the spatial pattern of the aggregate of TUBB2B and ACTG2 with their predicted off-target protein-coding genes (TUBB2A and ACTB/ACTA1/POTEM respectively) in Visium more closely resembled the spatial pattern of TUBB2B and ACTG2 observed in Xenium, accompanied by a corresponding decrease in RMSE for both genes and increase in Pearson correlation for TUBB2B. Likewise, a similar trend is observed in the scRNA-seq comparison (Supplementary Figure 8A-B). Ultimately, these findings suggest that off-target binding, even with imperfect sequence matching, can contribute to the expression patterns observed in Xenium.

RNA-seq reference atlases suggest off-target binding can variably impact results in Xenium custom probe panels

Our analyses to this point focused on the Xenium human breast gene panel from Janesick et al., a predecessor to and thus similar to the commercially available pre-designed Xenium v1 Human Breast Gene Expression Panel (Supplementary Note). We next sought to determine whether such off-target binding could affect custom Xenium gene panels, in which gene selection is specified by the user rather than provided as pre-designed panels by 10x Genomics. To investigate this, we leveraged two custom Xenium gene panels used by the Human BioMolecular Atlas Program (HuBMAP) (Jain et al. 2023): one designed for the placenta and another designed for the kidney, lung, and heart (i.e. multi-organ). Applying OPT to the associated FASTA files using the GENCODE v47 annotation and allowing for 10bp mismatches on either end of the 40bp probes, we found that 49 genes out of the 300 targeted genes in the placenta panel (Supplementary Table 8) and 24 genes out of the 300 targeted genes in the multi-organ (kidney, lung, heart) panel had predicted off-target binding (Supplementary Table 9). Of these, 30 of the 49 placenta panel genes and 11 of the 24 multi-organ panel genes had predicted off-targets that were protein-coding genes.

To assess the potential effects of our predicted off-target binding for these custom Xenium gene panels in experimental settings, we examined the expression of the corresponding predicted off-target genes in matched scRNA-seq or bulk RNA-seq data from the HuBMAP consortium for each of the relevant tissue types. For the placenta custom panel, 34 of the 49 genes with predicted off-target genes were detected with non-zero expression magnitudes in the placenta bulk RNA-seq dataset (Supplementary Figure 9). For the multi-organ panel, of the 21 genes with predicted off-target genes, 13 were detected with non-zero expression magnitudes in the heart, 12 in the kidney, and 13 in the lung (Supplementary Figure 10). Together, these results suggest that off-target binding can impact custom Xenium gene panels by distorting observed Xenium gene expression measurements in a tissue-dependent manner, particularly where the off-target gene is expressed at a higher magnitude compared to the target gene within the tissue of interest.

Discussion

Our study presents evidence of off-target probe binding that may distort gene expression profiles affecting the 10x Genomics Xenium spatial transcriptomics technology. We identified at least 14 out of the 313 genes in a Xenium human breast gene panel, which is highly similar to the commercially-available pre-designed Xenium v1 Human Breast Gene Expression Panel (Supplementary Note), that may be affected by off-target probe binding based on sequence similarity, supported by spatial and single-cell comparative analyses using Xenium with serial section datasets from Visium CytAssist and 3’ single-cell RNA-seq respectively. We further identified potential off-target probe binding affecting custom Xenium gene panels. To assist in the interpretation of existing probe-based gene expression data as well as future probe design, we provide OPT as a software tool for predicting potential off-target probe binding. For future reference, we have run OPT on all publicly available 10x Genomics pre-designed Xenium gene panels and supply them as a ZIP file (Supplementary Table 10).

Although we have predicted off-target binding based on sequence alignment, its effect on gene expression quantification may still vary. One reason is that the off-target protein-coding or non-protein-coding gene may not be expressed (Supplementary Figure 11). For example, in the Xenium human breast gene panel, although ADH1B probes have predicted off-target binding to ADH1A and ADH1C based on perfect sequence homology, the sparse expression of ADH1A and ADH1C in both the Visium and scRNA-seq breast cancer data led to only a minor difference in the aggregated expression and quantitative results (Supplementary Figure 12). Our analysis of HuBMAP custom probe panels demonstrated how to evaluate for the potential impact of predicted off-targets by integrating tissue-specific single-cell or bulk RNA-seq data from reference atlases (Supplementary Figures 9, 10). Overall, we anticipate evaluating whether predicted off-target genes are expressed in a tissue-specific manner will be useful for gauging whether predicted off-target binding is likely to meaningfully affect observed gene expression and interpretation when applied to a tissue of interest.

Other sources of non-specific signal may also arise, including probe self-hybridization or probe-probe interactions (Supplementary Fig 6B). In general, probe binding specificity is influenced by numerous factors, with many methods previously developed to aid in the design of probe sequences while taking these factors into consideration (Wang and Seed 2003; Rouillard et al. 2003; Wernersson and Nielsen 2005; Chou 2010; Li et al. 2011; Hu et al. 2020; Hershberg et al. 2021; Fornace et al. 2022; Stenberg et al. 2005; Kuemmerle et al. 2024). For example, in the Xenium human breast gene panel, HDC, a custom gene not included in the pre-designed Xenium v1 Human Breast Gene Expression Panel, did not have any off-targets predicted by OPT. Yet in Xenium, HDC exhibited a distinct spatial pattern and high global expression level, whereas in both Visium and scRNA-seq, HDC showed a minimal spatial pattern and sparse expression level respectively (Supplementary Figure 13). This illustrates how discrepancies across platforms can signal potential off-target activity not captured by alignment-based predictions alone and highlights the general importance of experimental validation with orthogonal technologies since sequence alignment-based tools such as OPT may not flag all potential discrepancies.

We note that most off-target binding impacts paralogs, homologous genes that have diverged following gene duplication events. Paralogs often belong to large gene families whose members can share high sequence similarity, increasing the risk of off-target probe activity. Unlike orthologs with conserved functions across species, paralogs are additional copies that can acquire function-altering mutations (Platt et al. 2000; Pevny et al. 1991; Tsai et al. 1994). Pooling expression signals across paralogs can therefore prevent researchers from capturing their distinct functional roles.

We also found that 10 of the 2,582 Xenium human breast gene panel probe sequences did not align to any reference transcripts in the GENCODE v47 annotation. Upon manually aligning these probes to the GRCh38 genome, we determined that each unmapped sequence corresponded either to regions immediately upstream or downstream of annotated transcripts, or to intronic sequences that would not typically be present in mature RNA. Interestingly, when we aligned several of these probes, such as (ENSG00000125878|TCF15|5d3cbc2) and (ENSG00000169083|AR|a0c6719), to an earlier annotation (GENCODE v28), they instead mapped to exonic regions (Supplementary Figure 14). This suggests that these intronic or intergenic probe sequences were likely designed using older GENCODE versions in which those regions were annotated as exons. This finding illustrates the importance of disclosing the specific annotation version to promote reproducibility, as well as the ongoing variability of human gene annotation. Likewise, as evidenced by our analysis across GENCODE, RefSeq, and CHESS, we emphasize the variation across these reference annotations and therefore recommend using multiple annotations when designing probe sequences and evaluating them for off-target effects to ensure a more comprehensive assessment.

Given these challenges, we advise probes with predicted off-target binding to protein-coding genes based on high sequence homology be avoided in future experiments. Likewise, we encourage the use of tools like OPT to aid in future probe design decisions and help ensure that probes are optimized to minimize off-target binding based on the most current transcriptome annotations. When probes with predicted off-targets cannot be avoided, we encourage the integration of tissue-specific RNA-seq data from HuBMAP and other reference atlases to evaluate for its potential impact. Further, such integration of tissue-specific RNA-seq data from reference atlases into the probe design process itself may offer a data-driven opportunity to minimize the impact of potential off-target binding by enforcing stricter probe-design constraints only where potential off-target genes are highly expressed in the tissue of interest. For datasets that have already been generated using probes with predicted off-target binding, we generally recommend taking into consideration these predictions to avoid drawing misleading conclusions. For example, we recommend expression measurements for genes with predicted off-target binding be omitted from training foundation models to avoid error propagation. Alternatively, when performing integrative analyses that compare or align gene expression with measurements across technologies, it may be necessary to incorporate off-target binding predictions. For instance, integration could be performed between the observed Xenium gene expression and the aggregated expression of the target and predicted off-target genes for the orthogonal technology. Finally, existing literature that base conclusions on genes with predicted off-target binding should be interpreted with caution.

We emphasize that these findings were missed in the previous Janesick et al. publication from 10x Genomics (Janesick et al. 2023). Consistent with previously published observations, we observed a highly correlated total gene expression magnitude between Xenium and Visium as well as scRNA-seq. However, a notable exception is APOBEC3B, which is not expressed according to both Visium and scRNA-seq but highly expressed according to Xenium (Figure 2B; Figure 3B) – a discrepancy that Janesick et al. omitted. We emphasize that positive significant average gene expression correlation is a necessary but not sufficient metric for consistency across technologies and that individual data points should be scrutinized. Likewise, validation with orthogonal technologies could have helped identify discrepancies suggestive of off-target effects. We note Janesick et al. used immunofluorescence to validate two genes, ERBB2 and MS4A1, which by our analysis were predicted to exhibit no off-target binding. Although Xenium incorporates blank and negative control probes that are intended to help quantify the rate of non-specific and potential off-target binding, our findings suggest that relying solely on such probes for error detection may be insufficient. Implementing probe redundancy, where the same gene is targeted using different codewords, could provide an additional internal control to enable the detection of off-target binding.

Although we focus here on the 10x Genomics Xenium technology, we do not exclude the possibility that off-target binding may similarly affect other probe-based gene detection approaches from other commercial vendors. Any technology that relies on hybridization-based detection is inherently susceptible to off-target probe binding when sequence similarity exists. Further, hybridization-based detection often inherently involves a trade-off between sensitivity and specificity. Given these inherent technological limitations, we therefore emphasize the importance of transparency through sharing probe sequences. However, many companies do not release the probe sequences used in their assays, limiting the consumer’s ability to fully interpret their results as well as the community’s ability to effectively characterize and benchmark performance variation across platforms. Therefore, we strongly recommend that companies publish probe sequences for pre-designed panels and likewise that researchers using these technologies should obtain and publish probe sequences used in their studies to support transparent and reproducible science.

This is not the first instance in which a commercially available platform has encountered challenges in probe design (McCartney et al. 2016; Harbig 2005; Mecham et al. 2004; Liu et al. 2014). These findings underscore the critical role of academic researchers towards ensuring the robustness of industry-led product development by providing oversight, free of financial conflicts of interest through independent federal funding. This complementarity between industry and academia fosters a more rigorous, transparent, and reliable scientific process, ultimately to the benefit of consumers and the public. By shedding light on putative off-target probe binding as well as by providing a tool to enable such off-target binding predictions, this work will help enhance the quality of spatial transcriptomics data and improve the overall reproducibility in spatial transcriptomics research.

Methods

Off-target Probe Tracker Tool (OPT)

OPT (Off-target Probe Tracker) is a Python program that runs nucmer (Marçais and Kingsford 2011) for alignment and then processes the results to predict probe binding based on sequence homology. OPT is available as an open-source Python toolkit at https://github.com/JEFworks-Lab/off-target-probe-tracker. When a user provides a query probe sequence file, a target transcript sequence file, and the annotation used to extract these transcripts, OPT outputs which gene each probe is likely to bind to. Nucmer is a fast nucleotide sequence aligner that uses maximal exact matches as anchors which it then joins together to find longer alignments. By default, OPT saves nucmer results in SAM format and finds perfect sequence matches between a query probe and a target transcript, requiring that alignments consist of only matches and cover the entirety of the query. OPT consists of four modules: (1) flip for reverse complementing probe sequences aligned to the opposite strand of their target genes; (2) track for aligning probe sequences and processing alignment results; (3) stat for compiling summary statistics on the number of off-target binding probes and affected genes; and (4) all for running the flip, track, and stat modules at once.

In the case that a probe’s target gene has synonyms, we consider alignments to genes annotated with one of its synonym to still be on-target. For example, if a probe that targets NARS shows alignments to a gene called NARS1, we don’t consider it to be off-target binding. We gathered relevant gene synonym relationships using the GeneCards and HGNC online database.

OPT also provides a “pad” mode in which imperfect alignments are allowed at either end of the query (i.e., probe sequence). The -pl parameter sets the pad length at either end of the query and OPT allows for any number of mismatches in these padded regions. For example, if the pad length is 10 and the probe sequence length is 40bp, then the middle 20bp are the only part of the probe sequence required to match. As long as the critical region is intact, OPT reports an off-target binding site based on this alignment. By default, -pl is set to 0, and the pad mode is activated by providing a non-zero integer to -pl.

Obtaining probe sequences for the Xenium v1 Human Breast Gene Expression Panel

To identify potential off-target binding impacting the 10x Genomics Xenium v1 Human Breast Gene Expression Panel, we obtained the FASTA file of probe sequences from the Janesick et al. publication courtesy of 10x Genomics and available as Supplementary Table 1 for preservation. Notably, this panel slightly deviates from the commercially available Xenium v1 Human Breast Gene Expression Panel (Supplementary Note).

The target gene names and IDs were extracted from the probe IDs of the following format:

> gene_id|gene_name|accession

We expected the provided FASTA file to contain probe sequences to be the reverse complemented sequence of their intended target genes and hence align to the reverse strand of their target isoforms. However, when we aligned the breast panel probe sequences to the GENCODE basic (v47) reference transcripts using nucmer, we found that 2,563 / 2,582 of probe sequences aligned on the reverse strand of their target transcripts (i.e., isoforms of their target genes). For consistency, we enforced that all probe sequences be oriented in the same direction and align to the forward strand of their target genes and transcripts. As such, we reverse-complemented these 2,563 probe sequences. We also added this functionality as an OPT module called “flip” in which probe sequences aligned to the reverse strand of their targets are reverse complemented. We expect probe sequences to align to the forward strand of transcripts (i.e., both oriented in the same direction) during the downstream probe sequence binding prediction step.

Visium Comparison

The Visium CytAssist dataset, collected from a breast cancer tissue block utilized in Janesick et al., was downloaded from the 10x Genomics website (https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast). This dataset originally contained 4,992 spots with x-y coordinates and included 18,085 genes per spot. Of the 313 unique genes in the Xenium dataset, 307 were shared with the Visium dataset; the other six genes (AKR1C1, ANGPT2, BTNL9, CD8B, POLR2J3, and TPSAB1) were excluded from the analysis because they were absent from the Visium dataset.

To compare spatial gene expression patterns from Visium and Xenium technologies, we first mapped all the data to the same coordinate space. We used STalign (v1.0.1), a computational tool that utilizes affine transformations along with diffeomorphic metric mapping to align target and source datasets (Clifton et al. 2023). The initial alignment involved only affine transformations and eight manually determined landmarks to align the Visium histology image (source) to the Xenium histology image (target). This transformation brought the Visium image into the coordinate space of the higher-resolution Xenium image. We then applied this learned transformation to the Visium spots, ensuring that they were correctly positioned relative to both histology images. Next, we used STalign to map the Xenium transcripts (source) onto their corresponding Xenium histology image (target) using both affine and diffeomorphic metric mapping. The transcripts were rasterized at 30μm resolution, with an initial affine transformation guided by four manually defined landmarks. Diffeomorphic metric mapping was then performed with the following parameters: a = 2500, epV = 1, niter = 2000, sigmaA = 0.11, sigmaB = 0.10, sigmaM = 0.15, sigmaP = 50, muA = [1, 1, 1], muB = [0, 0, 0], with all other settings left at their defaults. We extracted the overlapping regions between the two datasets (Supplementary Figure 3A), which reduced the total spots in the Visium dataset to 3,958. Finally, we aggregated the Xenium gene expression data to ∼55μm x 55μm patches that correspond to the spatial locations of the Visium spots, resulting in matched-resolution spatial gene expression for both technologies (Supplementary Figure 3B). The Visium spatial gene expression data is displayed as patches rather than spots to enhance visual saliency and ensure consistency with the Xenium spatial gene expression plots.

After obtaining matched-resolution spatial gene expression matrices, we quantified agreement between the Visium and Xenium data. For each gene shared between the two platforms, we constructed expression vectors across the aligned spatial spots, where each element corresponded to the log-normalized gene expression at its matched location in the tissue section. We then compared the Visium and Xenium vectors for each gene using two metrics: root-mean-square error (RMSE), computed relative to the line y = x, and Pearson correlation (r). To assess whether discrepancies in Xenium measurements could be attributed to predicted off-target genes, we compared the Xenium data of a target gene to the aggregated Visium expression of the target genes with its predicted off-targets. We summed the raw Visium counts of all off-target genes associated with each Xenium gene with predicted off-targets, re-normalized the aggregated counts using CPM followed by log(x + 1), and again calculated the RMSE and Pearson correlation. This approach enabled us to evaluate whether Xenium expression patterns aligned more closely with the intended target gene alone or with the combined expression of its predicted off-target genes.

Single-cell RNA-seq Comparison

The Chromium Next GEM 3’ scRNA-seq dataset, collected from a breast cancer tissue block utilized in Janesick et al., was downloaded from the 10x Genomics website (https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast). This dataset contained 12,388 cells with 36,601 genes per cell. All 313 unique genes present in the Xenium dataset are also in the scRNA-seq dataset, hence both datasets were subsetted to these genes for the analysis.

Both scRNA-seq and Xenium provide single-cell resolution data. To integrate these datasets, we first removed cells lacking detectable gene expression. We then normalized the combined gene expression data using counts per million (CPM) and applied a log transformation with a pseudocount of 1. Principal component analysis (PCA) is then applied to the normalized data, and batch effects are corrected using Harmony (v1.2.3) on the top 30 principal components (PCs) using default parameters except for theta, which was set to 8, to promote further mixing with clusters across technologies. Finally, Uniform Manifold Approximation and Projection (UMAP) is performed on the harmonized PCs, generating a shared 2D embedding across the two technologies, and the data is further facetted by technology for visualization (Supplementary Figure 5).

After generating a shared embedding, we quantified differences in gene expression patterns between scRNA-seq and Xenium. We first computed Leiden clusters on the harmonized principal components (resolution = 1.0) to identify transcriptionally similar groups of cells shared across both technologies. For each Leiden cluster, we calculated the mean expression of every gene present in both datasets. Clusters containing fewer than ten cells from either modality were excluded to ensure robust gene-level estimates. To compare expression patterns between scRNA-seq and Xenium for a given gene, we constructed vectors of cluster-level mean expression from both technologies and evaluated their similarity using two metrics: root-mean-square error (RMSE), computed relative to the line y = x, and Pearson correlation (r). To investigate whether predicted off-target genes may contribute to the observed target-gene expression in Xenium, we compared the Xenium target gene to the aggregated scRNA-seq expression of itself and its predicted off-targets. We summed the raw counts of a gene and all of its predicted off-targets in the scRNA-seq dataset, re-normalized the aggregated counts using CPM followed by log(x + 1), and again calculated the RMSE and Pearson correlation. This allowed us to test whether the Xenium gene expression more closely reflected the intended gene alone or the combined expression of the intended gene and its predicted off-targets.

Obtaining custom probe panels from HuBMAP

To evaluate potential off-target binding in the HuBMAP placenta and multi-tissue (heart, kidney, and lung) custom probe panels, we first downloaded the corresponding BED files from the HuBMAP portal (https://portal.hubmapconsortium.org/browse/dataset/28fe8e4ac8a4193f82fdd9f4d4eb0bb2; https://portal.hubmapconsortium.org/browse/dataset/6f597ca43db80f2499443f5c5bfac97c). Using pyfaidx and pandas in Python, we extracted each probe’s target gene name, gene identifier, and genomic coordinates from the BED files, and then generated FASTA files by retrieving the corresponding sequences from the reference genome (GRCh38). These FASTA files were then used as input to OPT to predict potential off-target binding.

HuBMAP custom probe-panel evaluation

To assess whether predicted off-target genes were likely to confound Xenium results in the HuBMAP custom probe panels, we evaluated the expression of the predicted off-target genes in matched HuBMAP RNA-seq datasets corresponding to the tissues for which each panel was designed. Four RNA-seq datasets were downloaded from the HuBMAP portal: a bulk RNA-seq dataset for the placenta (https://portal.hubmapconsortium.org/browse/dataset/cad68e5e910ca90bda79cc0d6aa586d1) and scRNA-seq datasets for the heart (https://portal.hubmapconsortium.org/browse/dataset/e62b7f592efe97894de86cfabbfddde8), kidney (https://portal.hubmapconsortium.org/browse/dataset/43f83f08bedac2bc4d58767362dfebe2), and lung (https://portal.hubmapconsortium.org/browse/dataset/6c2b7f1b8bc7e4fb6a1cadf29b09b83a).

For each tissue, raw count matrices were normalized to counts per million (CPM) and averaged across all cells to obtain a bulk-like mean expression profile. Log1p-transformed mean CPM values were used for all downstream comparisons. OPT was run on the two HuBMAP custom probe panels with all RNA species included and a pad length of 10. Since OPT reports transcript-level identifiers, we removed transcript-specific suffixes from Ensembl IDs to align them with gene symbols present in the RNA-seq datasets. For every target gene in the custom panels, we then compiled its predicted off-target genes based on OPT results and evaluated whether these off-targets were expressed in the matched tissue’s RNA-seq profile. To visualize these results, we generated heatmaps in which rows correspond to intended target genes and columns represent their predicted off-targets ordered by decreasing expression. This enabled direct comparison of the magnitude and tissue specificity of potential off-target expression across the HuBMAP datasets.

Cross-annotation analysis

To compare OPT’s results with different reference annotations, we used the most recent releases of GENCODE basic (v47), GENCODE comprehensive (v47), RefSeq (v110), and CHESS (v3.1.3) annotation of the GRCh38 genome. Note that GENCODE “basic” is the more reliable version of the annotation and is much closer to RefSeq and CHESS. GENCODE “comprehensive” includes hundreds of thousands of low-quality annotations, which we included in some of our analyses for completeness. Note also that GRCh38 has many non-reference sequences called “alternative scaffolds”; we removed these for our analysis. We then used gffread to extract transcripts as defined in these annotations by running:

$ gffread -w transcripts.fa -g grch38.p12/14.fa annotation.gff

The GRCh38.p14 assembly was used during transcript sequence extraction for all reference annotations, except for CHESS which specifies that the annotation maps genes and transcripts onto the GRCh38.p12 assembly. For RefSeq, we renamed the VD(J) segment features as transcript features to ensure consistency, and we also removed transcript sequences with the gene_biotype “pseudogene.” RefSeq has a separate biotype called “transcribed_pseudogene,” but doesn’t annotate transcripts for these features. We considered transcripts annotated for a small subset of just pseudogenes an error in the annotation.